---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "README-"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
library(hablar)
mtcars <- as_tibble(mtcars)
```

# hablar

hablar give users an easy and effective way to work with data types. Additionally, it provides non-astonishing results when summarizing data.

The ambition is to lower the barrier to R but also provides simple tools that experienced R users could benefit from.

## Installation

You can install hablar from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("davidsjoberg/hablar")
```

## convert

The most useful function of hablar is maybe convert. convert helps the user to quickly change data type of columns in a data frame. convert always converts factors to character before further conversion.

```{r example1}

## convert column:
# gear, vs to integers (int)
# cyl to factor (fct)
# am, gear and carb to character (chr)
mtcars %>% 
  convert(int(gear, vs),
          fct(cyl),
          chr(am, gear, carb))
```

For more information type `vignette("convert")` in the console.

## retype

A function for quick and dirty data type conversion. All columns are evaluated and converted to the simplest possible without loosing any information.

```{r example2}
## convert all columns to character
df <- mtcars %>% convert(chr(everything()))
df

## let retype guess the best data type
df %>% retype()
```

For more information type `vignette("retype")` in the console.

## s

Often summary function like min, max and mean returt suprising results. Combining s with your summary function ensures that you will get a result, if there is one in your data.

```{r example3}
## Base R
x <- c(NaN, 1, 2, NA)
min(x)

## With s
min(s(x))
```

For more information type `vignette("s")` in the console.

### Note
Hablar means 'speak R' in spanish. 