---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/hablar)](https://CRAN.R-project.org/package=hablar)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/hablar)
[![R-CMD-check](https://github.com/davidsjoberg/hablar/workflows/R-CMD-check/badge.svg)](https://github.com/davidsjoberg/hablar/actions)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/davidsjoberg/hablar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/davidsjoberg/hablar/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "README-"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
library(hablar)
mtcars <- as_tibble(mtcars)
```

# hablar

The mission of `hablar` is for you to get non-astonishing results! That means that functions return what you expected. R has some intuitive quirks that beginners and experienced programmers fail to identify. Some of the first weird features of R that `hablar` solves:

* Missing values `NA` and irrational values `Inf`, `NaN` is dominant. For example, in R `sum(c(1, 2, NA))` is `NA` and not 3. In `hablar` the addition of an underscore `sum_(c(1, 2, NA))` returns 3, as is often expected.

* Factors (categorical variables) that are converted to numeric returns the number of the category rather than the value. In `hablar` the `convert()` function always changes the type of the values.

* Finding duplicates, and rows with `NA` can be cumbersome. The functions `find_duplicates()` and `find_na()` make it easy to find where the data frame needs to be fixed. When the issues are found the utility replacement functions, e.g. `if_else_()`, `if_na()`, `zero_if()` easily fixes many of the most common problems you face.

`hablar` follows the syntax API of `tidyverse` and works seamlessly with `dplyr` and `tidyselect`. 

## Installation

You can install `hablar` from CRAN:

```{r cran_installation, eval = FALSE}
install.packages("hablar")
```

Or preferably:

```{r gh_installation, eval = FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, hablar)
```


## convert

The most useful function of `hablar` is maybe convert. convert helps the user to quickly and dynamically change data type of columns in a data frame. convert always converts factors to character before further conversion. Works with `tidyselect`.


```{r example1}
mtcars %>% 
  convert(int(cyl, am),
          fct(disp:drat),
          chr(contains("w")))
```

For more information type `vignette("convert")` in the console.

## Non-Astonishing summary functions

Often summary function like min, max and mean return surprising results. Combining `_` with your summary function ensures you that you will get a result, if there is one in your data. It ignores irrational numbers like `Inf` and `NaN` as well as `NA`. If all elements are `NA, Inf, NaN` it returns NA.

```{r example2}
starwars %>% 
  summarise(min_height_baseR = min(height),
            min_height_hablar = min_(height))
```

The function `min_` omitted that the variable `height` contained `NA`. For more information type `vignette("s")` in the console.

## Find the problem

When cleaning data you spend a lot of time understanding your data. Sometimes you get more row than you expected when doing a `left_join()`. Or you did not know that certain column contained missing values `NA` or irrational values like `Inf` or `NaN`.

In `hablar` the `find_*` functions speeds up your search for the problem. To find duplicated rows you simply `df %>% find_duplicates()`. You can also find duplicates in in specific columns, which can be useful before joins.

```{r}
# Create df with duplicates
df <- mtcars %>% 
  bind_rows(mtcars %>% slice(1, 5, 9))

# Return rows with duplicates in cyl and am
df %>% 
  find_duplicates(cyl, am)
```

There are also find functions for other cases. For example `find_na()` returns rows with missing values.

```{r}
starwars %>% 
  find_na(height)
```

If you rather want a Boolean value instead then e.g. `check_duplicates()` returns `TRUE` if the data frame contains duplicates, otherwise it returns `FALSE`.

##### **...apply the solution**

Let's say that we have found a problem is caused by missing values in the column `height` and you want to replace all missing values with the integer 100. `hablar` comes with an additional ways of doing if-or-else. 

```{r}
starwars %>% 
  find_na(height) %>% 
  mutate(height = if_na(height, 100L))
```

In the chunk above we successfully replaced all missing heights with the integer 100. `hablar` also contain the self explained:

* `if_zero()` and `zero_if()`
* `if_inf()` and `inf_if()`
* `if_nan()` and `nan_if()`

which works in the same way as the examples above.

## retype

A function for quick and dirty data type conversion. All columns are evaluated and converted to the simplest possible without loosing any information.

```{r example3}
mtcars %>% retype()
```

All variables with only integer were converted to type integer. For more information type `vignette("retype")` in the console.


### Note
Hablar means 'speak R' in Spanish. 
