---
title: "`retype` - auto data type conversion"
author: "David Sjoberg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{retype}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = F,
  comment = "#>"
)
library(DiagrammeR)
library(hablar)
library(dplyr)
```

## Can the data be simpler?
### `retype` quick start your analysis
Getting data into R can be hassle. But once you do, it often have incorrect data types/classes. It is not uncommon that numeric columns are characters or that dates are characters. Logical values could just as likely be 1 or 0. 

It is not uncommon that you realize a few hundred rows of code later that some column have the wrong data type, which could be annoying. It could also be a dangerous practise to quickly add `as.numeric` a few hundred lines after you import the data. The `hablar` package recommends that you correct all data types directly after you import the data into R! 

### What does `retype` do?
`retype` provides an easy approach for quick and dirty data type conversion. It follows a strict simplification hiarchy on each column of your data frame. It only converts the column if it assumes no important information is lost in the process. For example, the character vector `c("1", "2")` should rather be an integer vector. Likewise, the character `"2010-06-04"` should be a date. Factors have advantages, but they are never the simplest solution and hence it is always converted to character, at least.

### Usage
`retype(x, ...)`

where `x` is a data frame, and `...` is the column names you wanna apply `retype` to. `x` could also be a single vector.

### Simple example: numeric
```{r}
x <- as.numeric(3)
retype(x)
```
```{r}
class(retype(x))
```

### Simple example: character
```{r}
x <- as.character("2017-03-02")
retype(x)
```
```{r}
class(retype(x))
```

### Simple example: character
```{r}
x <- as.character(c("3,56", "0,78"))
retype(x)
```
```{r}
class(retype(x))
```

### Simple example: factor
```{r}
x <- as.factor(c(3, 4))
retype(x)
```
```{r}
class(retype(x))
```


## The simplification hiarchy
### Some things are simpler than others
`retype` uses a procedure to determine which data type is the simplest, without loosing any vital information in your data.

* The first thing to know about `retype` is that it always converts factors to character. 

* The second thing to know is that all logical columns are converted to integers.

* Thirdly, complex and list columns are left unchanged.

* From there it will test if it could be numeric vector. If true it converts the column to numeric.

* If it is numeric it tests if it could be an integer instead. If true, it converts the column to integer.

* If it is a character it tests if it could be a data time column. If true, it converts it to a date time (POSIXct) column.

* If it is a date time column it tests if it could be a date. If true, it converts it to a date column.


### A visualization of the hiarchy
The above procedure could be decripbed in a diagram. The arrows imply a test if a column could be converted to another without loosing information in your data. The procedure continues until it cannot be simplified further.

```{r, echo=F}
grViz("digraph d {

      node [shape = circle, style = filled]
      factor;character;numeric;integer;'date time';date;logical;list;complex

      logical -> integer;
      factor -> character;
      character -> numeric
      numeric -> integer;
      character -> 'date time';
      'date time' -> date;
      }")
```

## Example on a data frame
Examine the following dataset `starwars` from the package `dplyr`. First, we use `convert` on some columns to new data types.
```{r}
df <- starwars %>% 
  select(1:4) %>% 
  convert(fct(name),
           chr(height:mass),
           fct(hair_color)) %>% 
  print()
  
```
We then apply `retype` on `df`:
```{r}
df %>% 
  retype()
```

Which correctly guessed that height should be an integer and mass a numeric column. Factors were returned to character columns.

## Final notes
### `retype` in production code
Never use `retype` when you need your scripts to work the next time in the exact same way. `retype` may change over time, it could guess wrong and your data may change. Use `hablar::convert` instead where you explicitly state which data type each column should have.
