---
title: "`s` - short and simple summarizing"
author: "David Sjoberg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{s}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = F,
  comment = "#>"
)
library(dplyr)
library(hablar)
```

## Overview

### Description
The `s` function is a simple function that helps you get intuitive results when summarizing data. It is made to be used in conjuction with summarize functions, for example `min` , `sum` and `mean`. `s` takes a vector and mutates it in the following ways:

* It replaces all non-rational numbers from numeric vectors and replace them with `NA`. Non-rational numbers are `Inf`, `-Inf` and `NaN`.

* By default it removes `NA` from the vector

* If the vector has length zero or only consists of `NA` it returns a single `NA`.

### Usage

`s(..., ignore_na = T)`

where ... is one or more vector. If missing values should not be omitted use `ignore_na = F`.

### Simple examples

Removing `NA`:
```{r}
x <- c(NA, 1, 2)
s(x)
```

Replacing non-rational numbers with `NA` and then removes `NA`:
```{r}
x <- c(NaN, 1, Inf)
s(x)
```

Empty vectors return a single `NA`:
```{r}
x <- c()
s(x)
```

In conjuction with a summary function:
```{r}
x <- c(NaN, Inf, 3, 4)
median(s(x))
```

## The problems it solvs
### Principle of least astonishment
When programming the user hope that the output is intuitive. All programming languages have their special cases when you get astonishing results that you did not expect. The same goes for R. In the next parts of the `vignette` some problems it solves are explained in greater detail.

### Missing values
When learning R users might be surprised when creating suprised when using simple summary function. A summary function is a function that takes a vector and returns a single one value. For example, `min(x)` , `sum(x)` and `mean(x)`. A simple example:

```{r}
x <- c(1, 2, 3, 4, 5)
sum(x)
```

This simple example created the sum of 6 which is expected from adding one through five. However, there are many loopholes in R. New users to R might be confused that the next example results in `NA` (a missing value):

```{r}
x <- c(1, 2, 3, NA, 4)
mean(x)

```

Since the vector above have an a missing value R do not know how to find the mean of the vector. The missing value could be anything, and thus R returns the output `NA`. However, since missing values is a common reality when working with data it is a common practise to ignore missing values. Usually the user tells R to ignore the missing value and return the mean of the vector that have values that could be avaraged. The previous example could be fixed by adding `na.rm = TRUE` that drops all missing values before calculating the mean:

```{r}
x <- c(1, 2, 3, NA, 4)
mean(x, na.rm = TRUE)

```

Generally, R is strict about mssing values so that you do not miss them, which often is helpful rather than harsh! However, often the programmer want R to return a 'real' value from the data, if there is one, even if it ignores missing values.

The `s` function helps you with this. Since it by default removes missing values you can simply do:
```{r}
x <- c(1, 2, 3, NA, 4)
mean(s(x))

```

### Sometimes R removes too much

Adding an argument to remove all missing is common practise when summarizing data. However, it is not uncommon that some vectors only have missing values. Imagine an example where Amanda, David and Viktor sold sodas by the beach for three days. If someone did not show up they get a missing value.

```{r, echo=F}
df <- tibble(day = c(1, 2, 3, 1, 2, 3, 1, 2, 3),
       name = c(rep("Amanda", 3), rep("David", 3), rep("Viktor", 3)),
       sold_sodas = c(3, NA, 8, NA, NA, NA, 2, 1, 4)) %>% 
  print()
```

Now we want to see the most sodas each person could sell on a single day. The above data frame if saved as `df`.

```{r, warning=FALSE}
df %>% 
  group_by(name) %>% 
  summarize(n_sodas_best_day = max(sold_sodas, na.rm = T))
```

Amanda sold the most sodas in a single day. However, David that was abscent on all days got the result `-Inf`, meaning negative infinity as his best day. That is pretty astonishing! One could argue that it would be intuitive for him to get a missing value.

The reason for result is that we told R to remove all missing values before calculating the maximal value. It is equivalent to:

```{r, warning=FALSE}
x <- c()
max(x)
```

We could try to remove the `na.rm = TRUE` argument from `max()`.

```{r}
df %>% 
  group_by(name) %>% 
  summarize(n_sodas_best_day = max(sold_sodas))
```

Suddenly Viktor had the best day and Amanda, who was abscent the second day, got `NA` because R do not know how to find the maximal value. However, David also got `NA` this time, which makes sense. Working with data can be cumbersome! The `s` function is there to support you! Since it returns `NA` if the vector is empty we get:

```{r}
df %>% 
  group_by(name) %>% 
  summarize(n_sodas_best_day = max(s(sold_sodas)))
```


### Getting answers when there is none

Another astonishing result one might get is when R tries to return a value when there is none. Take this extract `df` from the `starwars` dataset from the R package `dplyr`. 

```{r, include = F}
df <- starwars %>% 
  select(name, homeworld, species, height)
```
```{r}
df %>% head(10)
```


Say that we want to calculate find the height of the tallest human from each homeworld. For safety, we drop all rows with missing values from the height column so that we do not get the same problem as before.

```{r, warning=FALSE}
df %>% 
  filter(!is.na(height)) %>% 
  group_by(homeworld) %>% 
  summarize(tallest_human = max(height[species == "Human"]))
```

We got negative infinity `-Inf` again. How could this be?

This is because some homeworld have no humans, e.g. Cerea. R tries to calculate the maximal value of nothing. The `s` function can help you out! Since it returns `NA` if the vector is empty we get:

```{r}
df %>% 
  filter(!is.na(height)) %>% 
  group_by(homeworld) %>% 
  summarize(tallest_human = max(s(height[species == "Human"])))
```

Now we get missing values for the homeworld that does not have any humans. Makes sense.



### Summarizing weird numbers

Numerical vectors in R can include more than numbers and missing values `NA`. They can aslo include infinite numbers `Inf` and `-Inf` as shown in the examples above. Furthermore, numerical vectors can include `NaN`'s which means 'not-a-number'. If the data frame you are using have `NaN` or `Inf` it may cause you problems when summarizing your data. Some examples:

```{r}
x <- c(NaN, 1)
min(x)
```

```{r}
x <- c(Inf, 3, 4)
mean(x)
```

```{r}
x <- c(5, -Inf, 2)
sum(x)
```

Often when you summmarize vectors that have `NaN` or `Inf` you want to treat them as a missing value. Maybe they have appeared as a mistake when you accidentally divided a value by zero since `1/0 = Inf` in R. The `s` function solves this for you be replacing them with `NA`.

```{r}
x <- c(NaN, 1)
min(s(x))
```

```{r}
x <- c(Inf, 3, 4)
mean(s(x))
```

```{r}
x <- c(5, -Inf, 2)
sum(s(x))
```

### Wrappers for `s` and summary functions

If things get too messy with an extra function you might prefer the wrapper functions of `s`. All major summary functions have an s wrapped alternative in `hablar`. These are names as the summary function added by an underscore, i.e. `min_(x)` and is equal to `min(s(x))`. The above examples with these wrappers for `s` are shown below.

```{r}
x <- c(NaN, 1)
min_(x)
```

```{r}
x <- c(Inf, 3, 4)
mean_(x)
```

```{r}
x <- c(5, -Inf, 2)
sum_(x)
```

## Final note

To summarize, `s` can help you to get results when you summarize your data, if there is an intuitive answer to be found, otherwise you will get `NA`.
